\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-07-11 tex/paper.tex
rewritten problem statement

In many fields of science such as particle physics, epidemiology,  and population
genetics, computer simulators are used to describe complex data generation processes. These simulators relate 
observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model. 
In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
% and may generate either discrete or continuous data. 
Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$. The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{abc, cranmer2015approximating}. 
Often the simulation involves control-flow that leads to models such that even $p(\bfx, \bfz | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.

In parallel, with the introduction of variational autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial networks~\cite{goodfellow2014generative}, there has been a vibrant research program around implicit generative models based on neural networks~\citep{2016arXiv161003483M}.  While these neural generative models also do not admit a tractable density, they are differentiable.

@@

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters $\bftheta$ of a non-differentiable, implicit
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
marginal distribution of generated data matches the observed data.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\bfz$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$, as
specified through a highly regularized deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the likelihood $p(\mathbf{x}|\bftheta)$ can be 
considered as a non-differentiable reparameterization of the random variable $\bfz$.
%rewritten as
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
%where $\mu$ is a probability measure.
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
e.g. as when $g$ is specified as a computer program.

>>>>

In many fields of science such as particle physics, epidemiology,  and population
genetics, computer simulators are used to describe complex data generation processes. These simulators relate 
observations $\bfx$ to the parameters $\bftheta$ of an underlying theory or mechanistic model. 
In most cases, these simulators are specified as procedural implementations of forward, stochastic processes involving latent variables $\bfz$.
% and may generate either discrete or continuous data. 
Rarely do these simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$. The prevalence and significance of this problem has motivated an active research effort in so-called \textit{likelihood-free inference} algorithms such as Approximate Bayesian Computation (ABC)~\cite{abc, cranmer2015approximating}. 
Often the simulation code involves control-flow that implies the dependence of $\bfx$ on $\bfz$ is non-differentiable, making inference even more difficult and precludes approaches such as Ref.~\citep{2016arXiv160507826G}.

%Often the simulation code involves control-flow that leads to models such that even $p(\bfx | \bftheta)$ is non-differentiable with respect to $\bftheta$, making inference even more difficult and precluding approaches such as Ref.~\citep{2016arXiv160507826G}.

In parallel, with the introduction of variational autoencoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial networks~\cite{goodfellow2014generative}, there has been a vibrant research program around implicit generative models based on neural networks~\citep{2016arXiv161003483M}.  While these implicit generative networks also do not admit a tractable density, they are differentiable.

@@

In this note, we develop an unsupervised learning algorithm for the point
estimation of the parameters $\bftheta$ of a non-differentiable, implicit
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.

%In this note, we develop a likelihood-free inference algorithm for the point
%estimation of the parameters $\bftheta$ of a non-differentiable, implicit
%generative model. We adapt the adversarial
%training procedure of generative adversarial
%networks~\cite{goodfellow2014generative} by replacing the implicit generative
%network with a domain-based scientific simulator, and solve the resulting
%non-differentiable minimax problem by minimizing variational upper
%bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
%objectives. The procedure results in learning a proposal distribution over
%simulator parameters, hence producing an arbitrarily tight family of models whose
%marginal distribution of generated data matches the observed data.

@@

We consider a family of parameterized densities $p(\mathbf{x}|\bftheta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\bftheta$ are the
parameters of interest. The simulation may involve some complicated latent
process
%, such that
%\begin{equation}\label{eqn:p_x}
%    p(\mathbf{x}|\bftheta) = \int p(\mathbf{x}|\bfz,\bftheta) p(\bfz|\bftheta) d\bfz
%\end{equation}
%\begin{equation}\label{eqn:p_x_sim}
%    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
%\end{equation}
where $\bfz \in {\cal Z}$ is a latent variable providing an external
source of randomness. 
Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
be a fixed-size vector with a simple density. Instead, the dimensionality of $z$ and the nature of it's components (uniform, normal, discrete, continuous, etc.) are inherited from the control flow of the simulation code and may depend on $\bftheta$ in some intricate way. Moreover, the dimensionality of $\bfz$ may be much larger than the dimensionality of $\bfx$.

%Unlike implicit generative models defined by neural networks, we do not assume $\bfz$ to
%be a fixed-size vector (e.g., it can be a sequence of variable length) and its
%distribution $p(\bfz|\bftheta)$ may itself depend on $\bftheta$ in some intricate way
%inherited from the control flow of the simulation code.

We assume that the stochastic generative process that defines $p(\mathbf{x}|\bfz,\bftheta)$ is
specified through a deterministic function $g(\cdot; \bftheta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\bftheta) \equiv \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta)
\end{equation}
such that the likelihood $p(\mathbf{x}|\bftheta)$ can be 
considered as a non-differentiable reparameterization of the random variable $\bfz$
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\bftheta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\bfz:g(\bfz;\bftheta) \leq \mathbf{x}\}} p(\bfz|\bftheta) \mu(d\bfz),
\end{equation}
where $\mu$ is a probability measure.
%Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
%used to generate data in forward mode. For this reason, evaluating the integral
%in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
%in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\bftheta$,
%e.g. as when $g$ is specified as a computer program.
